R for Clinical Study Reports and Submission

### 2021-09-19 project-folder.Rmd
add project folder information

> Here the `adam\` folder contain ADaM datasets. 
> The `tlf` folder contain output TLFs in `rtf` and `pdf` format.
> We put `adam\` and `tlf\` folder within `vignettes` folder only for illustration purpose. 
> In a real A&R report, you may have different location to save your input and output.

```
vignettes
├── adam
├── tlf
├── tlf-01-disposition.Rmd
├── tlf-02-population.Rmd
├── tlf-03-baseline.Rmd
├── tlf-04-efficacy.Rmd
├── tlf-05-ae-summary.Rmd
└── tlf-06-ae-spec.Rmd
```

@@

Reproducibility of analysis can be challenge. 
By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
we can easily achieve reproducibility for R and R package versions.

@@

In `esubdemo` project, a reproducible environment is created when you open the 
`esubdemo.Rproj` from RStudio. 
While RStudio open `esubdemo` project, RStudio will automatically execute R code in `.Rprofile`.
So we are using `.Rprofile` to setup a reproducible environment. 
More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html).

@@

After the `esubdemo` project is open, the code in `.Rprofile` will automatically check 
current R version is the same as we defined in `.Rprofile`

```{r, eval = FALSE}
R_version <- "4.0.1"	# set up project R version
```

@@

```
Error: The current R version is not the same as the current project in R4.0.1
```

If we are running under the same R version, 
you shall be able to see similar messages to inform how we control R package version.

@@

The R package version is controlled in two layers. 
Firstly, we define a snapshot date in `.Rprofile`. 
The snapshot date allows to freeze the repository. 
In other words, all R packages installed in this RStudio project 
are based on the frozen R version at the snapshot date. 
Here it is `2021-08-06` by using Microsoft R Application Network (MRAN) server.

```{r, eval = FALSE}
snapshot  <- "2021-08-06" 									                           # set up snapshot date
repos     <- paste0("https://mran.microsoft.com/snapshot/", snapshot)  # set up repository based on snapshot
```

Secondly, we use `renv` to lock R package versions and saved in `renv.lock` file. 
The exact same R package versions can be restored with `renv::restore()`.
An introduction of `renv` can be found in its [package website](https://rstudio.github.io/renv/articles/renv.html).

```
* Project '~/github-repo/esubdemo' loaded. [renv 0.14.0]
```

@@

For an organization, a cloud based R environment (e.g. RStudio Workbench) 
can enhance the development and operation of a clinical development team by 
centralize managing R and R package version. 
More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).

In summary, to achieve reproducibility for a project specific R package, 
a clinical project team can work under a controlled R environment 
in the same R version and R package versions defined by a repository snapshot date.

@@

By using the R package folder structure, you can also benefit from many outstanding tools to 
simplify and streamline your workflow.

There are additional folders and files required in this demo for an analysis project

> people may use different folder name as it is not a standard R package folder.
> These folders and files needs to be added in `.Rbuildignore` to pass
> R package check

- `adam`: ADaM datasets in `.sas7bdat` format.
  + one may also put it in `inst/exdata` following [R package convention](https://r-pkgs.org/data.html)
  + in reality, we suggest to have real data saved outside of this project. (e.g. in a database)
- `output`: TLFs output 
- `renv.lock` and `renv`: R package management using `renv` package. ([Introduction](https://rstudio.github.io/renv/articles/renv.html)) 
- `_pkgdown.yml`: [pkgdown](https://pkgdown.r-lib.org/articles/pkgdown.html) configuration file
- `.Rprofile`: Project startup file to setup running environment including R version, repository, folder path etc. 
  - We further use `inst/startup.R` and `R/zzz.R` to allow the startup file is executed while running. `devtools::load_all()` and RStudio build panel.

>>>>

> Here the `adam\` folder contain ADaM datasets. 
> The `tlf` folder contain output TLFs in RTF format.
> We put `adam\` and `tlf\` folder within `vignettes` folder only for illustration purpose. 
> In a real A&R report, you may have different location to save your input and output.

```
vignettes
├── adam_data
├── tlf
├── tlf-01-disposition.Rmd
├── tlf-02-population.Rmd
├── tlf-03-baseline.Rmd
├── tlf-04-efficacy.Rmd
├── tlf-05-ae-summary.Rmd
└── tlf-06-ae-spec.Rmd
```

@@

Reproducibility of analysis can be challenge. 
By using R package folder structure and proper tools (e.g. `renv`, `packrat`), 
we illustrate how to achieve reproducibility for R and R package versions.

@@

In `esubdemo` project, a reproducible environment is created when you open the 
`esubdemo.Rproj` from RStudio. 
While RStudio open `esubdemo` project, RStudio will automatically execute R code in `.Rprofile`.
So we can use `.Rprofile` to setup a reproducible environment. 
More details can be found in [https://rstats.wtf/r-startup.html](https://rstats.wtf/r-startup.html).

@@

In `esubdemo`, the `.Rprofile` contain two lines of R code.

```{r}
source("inst/startup.R")
source("renv/activate.R")
```

The first line execute R code saved in `inst/startup.R`. 
People can define a customized startup file to check running environment.

For example, we can require R version to be `4.1.1` by using code below. 
```{r, eval = FALSE}
R_version <- "4.1.1"	# set up project R version

# Check R Version
if(paste(R.version$major, R.version$minor, sep = ".") != R_version){
 warning("The current R version is not the same with the current project in ", R_version)
}
```

@@

```
Error: The current R version is not the same as the current project in R4.1.1
```

We can also define the package repository to be a specific snapshot date.
For example, we used Microsoft R Application Network (MRAN) to define the snapshot date to be `2021-08-06`. 
The snapshot date freeze the R package repository. 
In other words, all R packages installed in this RStudio project 
are based on the frozen R version at the snapshot date. 
Here it is `2021-08-06` by using MRAN server.

```{r}
# set up snapshot date
snapshot  <- "2021-08-06"

# set up repository based on snapshot
repos     <- paste0("https://mran.microsoft.com/snapshot/", snapshot)

# Define repo URL for project specific package installation
options(repos = repos)
```

The second line execute R code in `renv/activate.R`. 
This is a default behavior by using the `renv` R package to manage snapshot 
of the R packages. 
The `renv` provide a robust and stable approach to manage R package version 
for a project specific R packages.
As a user, you can use `renv::init()`, `renv::snapshot()` and `renv::restore()` 
to initialize, save and restore R packages used for the current analysis project.

In the analysis project, the `renv` package will

- create a `renv.lock` file to save the state of package versions.
- create a `renv` folder to manage R packages for a project.

An introduction of `renv` can be found in its [package website](https://rstudio.github.io/renv/articles/renv.html).

In summary, the R package version is controlled in two layers.

- Define a snapshot date in `inst/startup.R`.
- Using `renv` to lock R versions within a project.

If the project is initiated properly  
you shall be able to see similar messages to inform how we control R package versions.

@@

@@

> A cloud based R environment (e.g. RStudio Workbench) 
> can enhance the reproducibility within an organization 
> by using the same operation system, R and R package versions for an A&R project. 
> More details can be found in [https://environments.rstudio.com/](https://environments.rstudio.com/).

> A container like [docker](https://github.com/rstudio/r-docker) can further enhance 
> the reproducibility across an organization at operation system level, but beyond the scope of this book.

In conclusion, to achieve reproducibility for a project specific R package, 
a clinical project team can work under a controlled R environment 
in the same R version and R package versions defined by a repository snapshot date.

@@

By using the R package folder structure, you will benefit from many outstanding tools to 
simplify and streamline your workflow.

Here is a short list of commonly used R packages

- [`devtools`](https://devtools.r-lib.org/): make package development easier. 
  + A good overview can be found in [Chapter 2 of the R packages](https://r-pkgs.org/whole-game.html) book  
- [`usethis`](https://usethis.r-lib.org/): automates repetitive tasks that arise during project setup and development.
- [`testthat`](https://testthat.r-lib.org/): streamline testing code.
  + A discussion of using the `testthat` for an A&R project can be found in (@madhutest).  
- [`pkgdown`](https://pkgdown.r-lib.org/): generate static html documentation for an R package
  + It also allow you to run all analysis code in batch.
